Draft Article

### 2019-08-22 article.tex
introductory section structure

The US government's tepid response to the OPM hack illustrates a universally understood but rarely acknowledged fact about the relationship between cybersecurity and espionage: when a cyber operation appears to fall within the bounds of a traditional intelligence operation, it will merit the same diplomatic consequences that a traditional intelligence operation would have. Typically those consequences are minimal, rarely amounting to more than a pat denunciation of the activity and the necessary countermeasures to end it. Most importantly, those consequences never rise to the level where they impose additional costs on the state that attempted the espionage in the first place; in the rare cases when costs \emph{are} imposed for a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.

At first glance this observation might seem almost tautological in its simplicity. Of course cyber-espionage is treated like traditional espionage, one could argue---they're both espionage. That formulation begs the question;  it doesn't explain how states determine what constitutes \enquote{traditional} espionage, what purpose that distinction servers, or whether it is prudent to continue making it.

Despite the magnitude of the OPM hack, American officials had a hard time seeing it as anything other than a well-executed bit of espionage. A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].} For activities that fall under the umbrella of intelligence-gathering, counterintelligence, not prosecutions, are deemed to be the appropriate response.\footcite[In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets.]{finklea_cyber_2015} Director of the CIA John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, committed a classic Kinsley gaffe—saying what every official knows to be true but would never admit out loud: \enquote{You have to kind of salute the Chinese for what they did. If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015}

>>>>

The US government's tepid response to the OPM hack illustrates a universally understood but rarely acknowledged fact about the relationship between cybersecurity and espionage: when a cyber operation appears to fall within the bounds of a traditional intelligence operation, it will merit the same diplomatic consequences that a traditional intelligence operation would have. This justification was explicitly deployed by senior US policymakers, albeit often through unofficial moments of candor. A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footcite[In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets.]{finklea_cyber_2015} For activities that fall under the umbrella of intelligence-gathering, counterintelligence, not prosecution, is deemed to be the appropriate response. Director of the CIA John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence said what every official knows to be true but would never admit out loud: \enquote{You have to kind of salute the Chinese for what they did. If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015} Therefore, to understand why policymakers treat cyber-espionage as \enquote{fair game}, one needs to understand why traditional espionage is treated that way.

This article seeks to establish a theoretical framework for understanding the role that espionage plays in international relations (IR) and the diplomatic consequences that are associated with it. Typically those consequences are minimal, rarely amounting to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it. Most importantly, those consequences never rise to the level where they impose additional costs on the state that attempted the espionage in the first place; in the rare cases when costs \emph{are} imposed for a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again. Espionage retains a special status within the international community: the activity itself is expressly forbidden, but every state tacitly acknowledges that allies and adversaries alike are guaranteed to engage in it.

Because the diplomatic protocols behind espionage are relatively understudied from a social science perspective, this article makes two main contributions to the academic literature on espionage. The first is that it advances a theory for why espionage continues to retain its liminal status in the international legal system, where it is disavowed but allowed. Public discourse on this subject often uses terms like \enquote{rules of the road} or \enquote{gentleman's agreement} to describe the understanding between governments that espionage should fall within certain parameters. By employing both realist and constructivist schools of IR, this article develops rigerous explanations for why rational states, whether they be security-seeking and greedy, might cooperate to preserve this dynamic.

\section{Existing research on espionage}
Previous scholarly work examines the curious liminal space in which espionage exists, but not the rationale that states have for perpetuating it.

\section{Cyber-espionage as espionage}
At first glance this observation might seem almost tautological. Of course cyber-espionage is treated like traditional espionage, one could argue---they're both espionage. That formulation begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage, what purpose that distinction servers, or whether it is prudent to continue making it.
